Hierarchical time series forecasting in Emergency Medical Services

### 2022-11-16 manuscript/03-experiment.qmd
render to pdf

We are interested in generating forecasts to inform the planning horizon of $ph= 42$ days, interested by planners in the ambulance services trust. The forecast horizon in this study is $fh = 2 \times ph$ days ahead (2*42 days planning horizon). This is because the planning is generally frozen for $ph$ days and considering a forecast horizon of $ph$ days might not be helpful for planning. While forecasts are generated for $2 \times ph$ days ahead, performance evaluation is only assessed based on the last $ph$ days and not the $2 \times ph$ days. The forecasts are produced for the holdout of 365 days using time series cross-validation [@hyndman2021forecasting].
In the following section, we discuss the dataset, describe the forecasting methods used to generate base forecasts and present the point and probabilistic accuracy measures.

@@

Given the total number of time series, direct visual analysis is infeasible. Therefore, we first compute features of all 1530 time series and display the strength of trend and weekly seasonality strength in Figure @fig-feature. Each point represents one time series with strength of trend in  x-axis and strength of seasonality in y-axis. It is clear that there are some series showing strong trends and/or seasonality, corresponding to series at the higher levels of the hierarchy. The majority show low trend and seasonality with . These are time series belonging to the bottom of the structure, series related to the nature of incidents for each given control, health board and priority.

```{r}
#| label: fig-feature
#| cache: true
#| out.width: "70%"
#| fig.align: center
#| fig-cap: "Time series features of attended incidents across all levels (1530 series)"
incident_gthf <- readr::read_rds(here::here("data/incidents_gt.rds"))
incident_gthf %>%
  features(incident, feat_stl) %>%
  ggplot(aes(x = trend_strength, y = seasonal_strength_week)) +
  geom_point(alpha = 0.5) +
  ggthemes::theme_few() +
  labs(x = "Strength of trend", y = "Strength of weekly seasonality")
```

In addition to displaying the trend and seasonality features, we also highlight few time series at various levels of the aggregation. Figure @fig-dataviz reveals different information such as trend, seasonality and noise. For example, some series depict seasonality and trend, whereas some other series report low volume of attended incidents and entropy, making them more volatile and difficult to forecast. At the level on nature of incidents combined with categories of other levels, there are many series that contain zeros with low counts. As such, the data set represents a diverse set of daily time series patterns.

@@

Forecasting performance is evaluated using both point and probabilistic error measures.

The point forecast accuracy is measured via Root Mean Squared Scaled Error (RMSSE):

$$
\text{RMSSE} = \sqrt{\text{mean}(q_{j}^2)},
$$ {#eq-RMSSE}

>>>>

We are interested in generating forecasts to inform the planning horizon of $ph= 42$ days, required by planners in the ambulance services trust. The forecast horizon in this study is $fh = 2 \times ph$ days ahead (2*42 days planning horizon). This is because the planning is generally freezed for $ph$ days and considering a forecast horizon of $ph$ days might not be helpful for planning. While forecasts are generated for $2 \times ph$ days ahead, performance evaluation is only assessed based on the last $ph$ days and not the $2 \times ph$ days. The forecasts are produced for the holdout of 365 days using time series cross-validation [@hyndman2021forecasting].

In the following section, we discuss the dataset, describe the forecasting methods used to generate base forecasts and present the point and probabilistic accuracy measures.

@@

Given the total number of time series, direct visual analysis is infeasible. Therefore, we first compute features of all 1530 time series and display the strength of trend and weekly seasonality strength in Figure @fig-feature. Each point represents one time series with the strength of trend in x-axis and the strength of seasonality in y-axis. It is clear that there are some series showing strong trends and/or seasonality, corresponding to series at the higher levels of the hierarchy. The majority of series show low trend and seasonality. These are time series belonging to the bottom series, series related to the nature of incidents for a given control, health board and priority level. Bottom series are dominated by noise with little or no systematic patterns.

```{r}
#| label: fig-feature
#| cache: true
#| out.width: "70%"
#| fig.align: center
#| fig-cap: "Time series features of attended incidents across all levels (1530 series)"
incident_gthf <- readr::read_rds(paste0(storage_folder, "incidents_gt.rds"))
incident_gthf %>%
  features(incident, feat_stl) %>%
  ggplot(aes(x = trend_strength, y = seasonal_strength_week)) +
  geom_point(alpha = 0.5) +
  ggthemes::theme_few() +
  labs(x = "Strength of trend", y = "Strength of weekly seasonality")
```

In addition to displaying the trend and seasonality features, we also visualize few time series at various levels of the aggregation. Figure @fig-dataviz reveals different information such as trend, seasonality and noise. For example, some series depict seasonality and trend, whereas some other series report low volume of attended incidents and entropy, making them more volatile and difficult to forecast. At the level on nature of incidents combined with categories of other levels, there are many series that contain zeros with low counts. As such, the data set represents a diverse set of daily time series patterns.

@@

The data is spitted into training (up until 2018-07-31) and test (from  2018-08-01 to the end) sets, with all model development and hyper-parameter tuning performed using training data only. The time series cross validation is performed with a forecast horizon of $2*42$ days and advances in 42-day steps. Forecasting performance is evaluated using both point and probabilistic error measures. Forecast error is calculated by considering only the last 42 days and not the entire $2*42$ days. This corresponds on how forecasts are generated for planning in practice. Forecasting performance is evaluated using both point and probabilistic error measures. The point forecast accuracy is measured via Root Mean Squared Scaled Error (RMSSE) and Mean Absolute Scaled Error (MASE) that are describes below:

$$
RMSSE = \sqrt{mean(q_{j}^2)},
$$ {#eq-RMSSE}

### 2022-11-16 manuscript/main.tex
render to pdf

\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[many]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\journal{Annals of Emergency Medicine}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{elsarticle-harv}
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Hierarchical time series forecasting in Emergency Medical Services},
  pdfauthor={Bahman Rostami-Tabar; Rob J. Hyndman},
  pdfkeywords={forecasting, healthcare, emergency services, forecast
reconciliation, hierarchical time series, ambulance demand, attended
incidents},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{keyword}
    forecasting \sep healthcare \sep emergency services \sep forecast
reconciliation \sep hierarchical time series \sep ambulance demand \sep 
    attended incidents
\end{keyword}
\end{frontmatter}
    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, breakable, interior hidden, frame hidden, borderline west={3pt}{0pt}{shadecolor}, sharp corners, boxrule=0pt]}{\end{tcolorbox}}\fi

@@

We are interested in generating forecasts to inform the planning horizon
of \(ph= 42\) days, interested by planners in the ambulance services
trust. The forecast horizon in this study is \(fh = 2 \times ph\) days
ahead (2*42 days planning horizon). This is because the planning is
generally frozen for \(ph\) days and considering a forecast horizon of
\(ph\) days might not be helpful for planning. While forecasts are
generated for \(2 \times ph\) days ahead, performance evaluation is only
assessed based on the last \(ph\) days and not the \(2 \times ph\) days.
The forecasts are produced for the holdout of 365 days using time series
cross-validation \citep{hyndman2021forecasting}. In the following
section, we discuss the dataset, describe the forecasting methods used
to generate base forecasts and present the point and probabilistic
accuracy measures.

@@

The dataset used in this study is from a major ambulance service trust
in the United Kingdom. It contains information relating to the daily
number of attended incidents from 10 October 2015 to 31 July 2019 by
nature of incidents, priority, the health board managing the service and
the control area (or region). Figure Figure~\ref{fig-hierarchy} depicts
both the hierarchical and grouped structure of the data. Figure
Figure~\ref{fig-hierarchy-1} illustrates the nested hierarchical
structure based on control area and health board and Figure
Figure~\ref{fig-hierarchy-2} shows the grouped structure by priority and
the nature of incident.

@@

\includegraphics{../img/group.png}

@@

Given the total number of time series, direct visual analysis is
infeasible. Therefore, we first compute features of all 1530 time series
and display the strength of trend and weekly seasonality strength in
Figure Figure~\ref{fig-feature}. Each point represents one time series
with strength of trend in x-axis and strength of seasonality in y-axis.
It is clear that there are some series showing strong trends and/or
seasonality, corresponding to series at the higher levels of the
hierarchy. The majority show low trend and seasonality with . These are
time series belonging to the bottom of the structure, series related to
the nature of incidents for each given control, health board and
priority.

@@

In addition to displaying the trend and seasonality features, we also
highlight few time series at various levels of the aggregation. Figure
Figure~\ref{fig-dataviz} reveals different information such as trend,
seasonality and noise. For example, some series depict seasonality and
trend, whereas some other series report low volume of attended incidents
and entropy, making them more volatile and difficult to forecast. At the
level on nature of incidents combined with categories of other levels,
there are many series that contain zeros with low counts. As such, the
data set represents a diverse set of daily time series patterns.

@@

\textbf{Naive:} We start with one of the simplest forecasting approaches
used in practice - assuming that in the next few days, everything will
be the same as in the similar days in the past. This is called
``Naïve''. In our case, given that we need a distribution of values, we
will use a modified approach, where the empirical distribution of the
daily attended incidents time series is used to forecast the future
attended incidents distribution. We consider the empirical distribution
of the most recent year of historic data on a rolling basis to capture
potential changes in behavior over time.

@@

\textbf{Poisson Regression: } Despite the popularity and the relevant of
automatic \texttt{ETS} and \texttt{ARIMA} in this study, but they
produce non-integer attended incidents and forecasts might also be
negative. However, the number of attended incidents is an integer and
non-negative. When using \texttt{ETS} and \texttt{ARIMA}, a time series
transformation approach such as \texttt{sqrt()} could be used to
generate strictly positive forecasts, however forecasts are still not
integer. Another alternative is to use forecasting models that produce
integer, non-negative forecasts. One of the models that is frequently
used in practice is the Poisson regression \citep[see, for
example,][]{mccarthy2008challenge}, which can be summarized as
\begin{equation}\protect\hypertarget{eq-poisson-regression}{}{
  \begin{aligned}
    & {y}_t \sim \mathrm{Poisson} \left( \lambda_t \right) \\
    & \log \lambda_t = \mathbf{x}_t' \boldsymbol{\beta}
  \end{aligned} \quad
}\label{eq-poisson-regression}\end{equation}

The logarithm in Equation~\ref{eq-poisson-regression} is needed to
ensure that the parameter of Poisson distribution is always positive.
This model can be estimated via maximization of the likelihood function
based on Poisson mass function. For taking into account serial
dependence we include a regression on the previous observation. We try
to capture the short range serial dependence by a first order
autoregressive term and the weekly seasonality by a 7th order
autoregressive term. We use the \texttt{tscount()} function in the fable
package in R, which is a wrapper function written based on
\texttt{tscount} package. We fit a model to each time series using the
function tsglm, with the identity link function, defined by the argument
link.

@@

Forecasting performance is evaluated using both point and probabilistic
error measures.

The point forecast accuracy is measured via Root Mean Squared Scaled
Error (RMSSE):

\begin{equation}\protect\hypertarget{eq-RMSSE}{}{
\text{RMSSE} = \sqrt{\text{mean}(q_{j}^2)},
}\label{eq-RMSSE}\end{equation}

@@

\[
q^2_{j} = \frac{\displaystyle e^2_{j}}
    {\displaystyle\frac{1}{T-m}\sum_{t=m+1}^T (y_{t}-y_{t-m})^2},
\]

\(e_{j}\) is the point forecast error \(j\) and \(m = 1\) for
non-seasonal series and \(m = 7\) for daily seasonal series, \(y_t\) is
the observation for period \(t\) and \(T\) is the sample size
(observations used for training the forecasting model). Smaller RMSSE
values suggest more accurate forecasts. Note that the measure is
scale-independent, thus allowing us to average the results across
series.

>>>>

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[many]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\journal{Annals of Emergency Medicine}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{elsarticle-harv}
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Hierarchical time series forecasting in Emergency Medical Services},
  pdfauthor={Bahman Rostami-Tabar; Rob J. Hyndman},
  pdfkeywords={forecasting, healthcare, emergency services, forecast
reconciliation, hierarchical time series, ambulance demand, attended
incidents},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{keyword}
    forecasting \sep healthcare \sep emergency services \sep forecast
reconciliation \sep hierarchical time series \sep ambulance demand \sep 
    attended incidents
\end{keyword}
\end{frontmatter}
    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, interior hidden, breakable, enhanced, frame hidden, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi

@@

We are interested in generating forecasts to inform the planning horizon
of \(ph= 42\) days, required by planners in the ambulance services
trust. The forecast horizon in this study is \(fh = 2 \times ph\) days
ahead (2*42 days planning horizon). This is because the planning is
generally freezed for \(ph\) days and considering a forecast horizon of
\(ph\) days might not be helpful for planning. While forecasts are
generated for \(2 \times ph\) days ahead, performance evaluation is only
assessed based on the last \(ph\) days and not the \(2 \times ph\) days.
The forecasts are produced for the holdout of 365 days using time series
cross-validation \citep{hyndman2021forecasting}.

In the following section, we discuss the dataset, describe the
forecasting methods used to generate base forecasts and present the
point and probabilistic accuracy measures.

@@

The dataset used in this study is from a major ambulance service trust
in Great Britain. It contains information relating to the daily number
of attended incidents from 10 October 2015 to 31 July 2019 by nature of
incidents, priority, the health board managing the service and the
control area (or region). Figure Figure~\ref{fig-hierarchy} depicts both
the hierarchical and grouped structure of the data. Figure
Figure~\ref{fig-hierarchy-1} illustrates the nested hierarchical
structure based on control area and health board and Figure
Figure~\ref{fig-hierarchy-2} shows the grouped structure by priority and
the nature of incident.

@@

\includegraphics{/Users/bahmanrostami-tabar/Documents/1-research/1-research_papers/Work with Rob Hyndman/forecasting-emergency-medcine/img/group.png}

@@

Given the total number of time series, direct visual analysis is
infeasible. Therefore, we first compute features of all 1530 time series
and display the strength of trend and weekly seasonality strength in
Figure Figure~\ref{fig-feature}. Each point represents one time series
with the strength of trend in x-axis and the strength of seasonality in
y-axis. It is clear that there are some series showing strong trends
and/or seasonality, corresponding to series at the higher levels of the
hierarchy. The majority of series show low trend and seasonality. These
are time series belonging to the bottom series, series related to the
nature of incidents for a given control, health board and priority
level. Bottom series are dominated by noise with little or no systematic
patterns.

@@

In addition to displaying the trend and seasonality features, we also
visualize few time series at various levels of the aggregation. Figure
Figure~\ref{fig-dataviz} reveals different information such as trend,
seasonality and noise. For example, some series depict seasonality and
trend, whereas some other series report low volume of attended incidents
and entropy, making them more volatile and difficult to forecast. At the
level on nature of incidents combined with categories of other levels,
there are many series that contain zeros with low counts. As such, the
data set represents a diverse set of daily time series patterns.

@@

\textbf{Naive:} We start with one of the simplest forecasting approaches
used in practice - assuming that in the next few days, everything will
be the same as similar days in the past. This is called ``Naïve''. In
our case, given that we need a distribution of values, we will use a
modified approach, where the empirical distribution of the daily
attended incidents time series is used to forecast the future attended
incidents distribution. We consider the empirical distribution of the
most recent year of historic data on a rolling basis to capture
potential changes in behavior over time.

@@

Despite the popularity and the relevance of automatic \texttt{ETS} in
this study, however it may produce forecasts that are non-integer and
negative. However, the number of attended incidents is an integer and
non-negative. When using \texttt{ETS} , a time series transformation
approach could be used to generate strictly positive forecasts, however
forecasts are still not integer. An alternative is to use forecasting
models that produce integer, non-negative forecasts. in the following
section we present Generalized Linear Models (GLMs) and Poisson
Regression to produce count base forecasts.

\textbf{Generalized Linear Model (GLM): } GLMs are a family of models
developed to extend the concept of linear regression models. They
perform a regression by modeling the response variable as coming from a
particular member of the exponential family, and then transforming the
mean of the response so that the transformed mean is a linear function
of the predictors. One of the models that is frequently used in practice
to generate count forecasts is the Poisson regression. We will consider
forecasting attended incidents using the covariates spline trend, day of
the week dummy variables (from Monday to Sunday), Fourier terms to
capture yearly seasonality, public holidays (1 when is public holiday, 0
otherwise), school holidays (1 when is school holiday, 0 otherwise) and
Christmas Day (1 when is Christmas Day, 0 otherwise) and New Year's Day
(1 when is New Year's Day, 0 otherwise). We fit a Poisson regression
model using the function \texttt{glm()} from the package in R, with the
argument \texttt{family\ =\ poisson} to specify that we wish to fit a
Poisson regression model.

\textbf{Poisson Regression using tscount: } We consider another Poisson
regression model that takes into account serial dependence in addition
to covariates used in the GLM model. To that end, we use
\texttt{tsglm()} function in \emph{tscount} package in R to model the
attended incidents. The logarithmic link function is used to ensure that
the parameter of Poisson distribution is always positive. This model can
be estimated via maximization of the likelihood function based on
Poisson mass function. The regression model captures the short range
serial dependence by including the three order autoregressive terms.

\textbf{Ensemble method:} we also use an ensemble method that combines
forecasts generated from Naive, ETS, GLM and Poisson regression together
to make more accurate forecasts than any individual model.

@@

The data is spitted into training (up until 2018-07-31) and test (from
2018-08-01 to the end) sets, with all model development and
hyper-parameter tuning performed using training data only. The time
series cross validation is performed with a forecast horizon of \(2*42\)
days and advances in 42-day steps. Forecasting performance is evaluated
using both point and probabilistic error measures. Forecast error is
calculated by considering only the last 42 days and not the entire
\(2*42\) days. This corresponds on how forecasts are generated for
planning in practice. Forecasting performance is evaluated using both
point and probabilistic error measures. The point forecast accuracy is
measured via Root Mean Squared Scaled Error (RMSSE) and Mean Absolute
Scaled Error (MASE) that are describes below:

\begin{equation}\protect\hypertarget{eq-RMSSE}{}{
RMSSE = \sqrt{mean(q_{j}^2)},
}\label{eq-RMSSE}\end{equation}

@@

\[
q^2_{j} = \frac{ e^2_{j}}
    {\frac{1}{T-m}\sum_{t=m+1}^T (y_{t}-y_{t-m})^2},
\] \(e_{j}\) is the point forecast error \(j\) and \(m = 1\) for
non-seasonal series and \(m = 7\) for daily seasonal series, \(y_t\) is
the observation for period \(t\) and \(T\) is the sample size
(observations used for training the forecasting model). Smaller RMSSE
values suggest more accurate forecasts. Note that the measure is
scale-independent, thus allowing us to average the results across
series.

Mean Absolute Scaled Error (MASE) is calculated as:

\[
MASE = mean(|q_{j}|),
\]

where, \[
q_{j} = \frac{ e_{j}}
    {\frac{1}{T-m}\sum_{t=m+1}^T |y_{t}-y_{t-m}|}.
\]

The denominator is the mean absolute error of the naive method in the
fitting sample of \(n\) observations and is used to scale the error.
